Type theories as quotient inductive-recursive types

### 2025-02-11 tex/main.tex
S2.3 on overlapping patterns

\supplement{The formal development is hosted at the \textsc{GitHub} repository: \url{https://github.com/genetsai95/DTT-QIIRT}.\LT[noinline]{Transfer it to my account?}}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
%
We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.
\LT{We write $\Set$ for the type of types employing the `typical ambiguity'}

@@

We have formally implemented our definitions and proofs in Agda, with the following options: \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}. We simulate quotient inductive types using postulated equations to introduce equality constructors~\cite{Licata2011}.
Some of our theorem statements are meta-theoretical, and thus cannot be checked by Agda. Others can only be checked  partially, because of postulated quotient inductive types. We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
\LT{Single quote or double quote in British English?}

%\LT{
%Points to include:
%\begin{enumerate}
%  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
%  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
%  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
%\end{enumerate}}
%
%\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive typeswbe checked ).}

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\cite{Cockx2014,Altenkirch2016a}
\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & \defeq A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & \defeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
where we have written $t \reduce u$ for $t \equiv u$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

To ensure that type substitution is strict for all %\emph
{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
This option is turned on globally in our formal implementation.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

  For example, $[\sigma;\tau]_{\Ty}\;\UU \reduce [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \reduce [\sigma]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-5} as the first reduction.
  It is easy to see that remaining local peaks reduce to the same term.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
\end{proof}

@@

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
\end{proof}

@@

\begin{proposition}[Local confluence]\label{prop:local-confluence-3}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in \cref{prop:local-confluence-1,prop:local-confluence-2}, consider
  \begin{align*}
    [\idS]_{\Ty}\;(\Pi\;A\;B) && [\sigma;\tau]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma, t)]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma; \tau)]_{\Ty}\;(\Pi\;A\;B).
  \end{align*}
  Each of them reduces to the following terms after applying clauses of $\blank\uparrow\blank$ and $[\blank]_{\Ty}\blank$:
  \begin{align*}
    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}\,[\tau \uparrow A ]_{\Ty}\,B) \\
    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}\,[\pi_1\tau \uparrow A]_{\Ty}\,B)
  \end{align*}
  respectively.
  Hence, the type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}

>>>>

\supplement{The formal development is hosted at the \textsc{GitHub} repository: \url{https://github.com/L-TChen/TTasQIIRT}}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
%
We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which is not given explicitly, but instead can be figured out from context.
We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.

@@

We have formally implemented our definitions and proofs in \Agda, using the options \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}.
Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}.
Some of our statements are meta-theoretical and thus not checked by \Agda.
Others are only partially checked, as \Agda does not generate proof obligations for postulated quotient inductive types directly.
We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
Functions are typically defined informally using pattern matching for convenience and readability. Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
This elaboration enables conventional notation while making computational behaviour opaque and favouring asymmetric definitions.
For example, the following two definitions are essentially the same

\begin{minipage}[t]{.5\textwidth}
  \begin{alignat*}{5}
    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}[t]{.5\textwidth}
\begin{alignat*}{5}
  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\,n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' m)
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
after elaboration in \Agda: each clause of $\blank +' \blank$ is strict (hence the symbol $\defeq$ is used) while clauses of $\blank+\blank$ are not necessarily strict by definition.
This semantics for function definitions affects how definitions reduce and therefore how identities are proved.

Cockx explores an alternative semantics for pattern matching~\cite{Cockx2014}, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
Under this semantics, each clause of $\blank + \blank$ is strictly equal to its right-hand side, allowing the symmetry of $\blank + \blank$ to be proved commutative directly by induction on $n$ and $m$.
Thus, when a function $f$ is defined with overlapping patterns, we use the notation $f\;\overrightarrow{p_i} \reduce t_i$ to emphasize its reduction behaviour.

A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any clause must eventually compute to the same term, i.e.\ $f$ is confluent.
Cockx shows~\cite[Proposition~2]{Cockx2014} that if a definition of $f$ is
\begin{enumerate}
  \item \emph{structurally recursive}, i.e.\ arguments of recursive calls are structurally smaller, and
  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, their right-hand sides $\sigma t_1$ and $\sigma t_2$ are strictly equal, 
\end{enumerate}
then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.

\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
where we have written $t \reduce u$ for $t \equiv u$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

To ensure that type substitution is strict for all %\emph
{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
This option is turned on globally in our formal implementation.}

\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

  For example, $[\sigma;\tau]_{\Ty}\;\UU \reduce [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \reduce [\sigma]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-5} as the first reduction.
  It is easy to see that remaining local peaks reduce to the same term.
\end{proof}

@@

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}

@@

\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proposition}[Local confluence]\label{prop:local-confluence-3}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in \cref{prop:local-confluence-1,prop:local-confluence-2}, consider
  \begin{align*}
    [\idS]_{\Ty}\;(\Pi\;A\;B) && [\sigma;\tau]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma, t)]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma; \tau)]_{\Ty}\;(\Pi\;A\;B).
  \end{align*}
  Each of them reduces to the following terms after applying clauses of $\blank\uparrow\blank$ and $[\blank]_{\Ty}\blank$:
  \begin{align*}
    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}\,[\tau \uparrow A ]_{\Ty}\,B) \\
    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}\,[\pi_1\tau \uparrow A]_{\Ty}\,B)
  \end{align*}
  respectively.
  Hence, the type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
\end{proof}
